Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-02 doc/types-pldi2013/evaluation.tex
end

We evaluate the design of Legion's static and dynamic semantics on four criteria:
expressivity (can real applications be written---Section~\ref{subsec:expressivity}),
overhead  (what are the dynamic checking costs---Section~\ref{subsec:overhead}),
scalability (can it enable hierarchical scheduling---Section~\ref{subsec:scalability}),
and performance  (does the performance increase from relaxed coherence modes 
warrant the increased semantic complexity---Section~\ref{subsec:relaxedperf}).
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

\subsection{Expressivity}
\label{subsec:expressivity}
We evaluate Legion on three real-world applications.  To qualitatively gauge the 
expressivity of the Legion type system, we introduce these applications
by describing features of the Legion type system used in their implementations.
The Circuit example was already covered in detail in Section~\ref{sec:example}.

@@

The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with sub-refinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every level of refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

@@

Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence 
are needed in some application;  region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multi-colorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications.
% written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

\subsection{Checking Overhead}
\label{subsec:overhead}
Our first Legion implementation consisted of a C++ library of Legion primitives
\cite{Legion12} with no checking of region memory accesses.  
When testing applications we frequently encountered memory corruption due to
illegal region accesses caused by application bugs.
In many cases, this corruption occurred between nodes in the cluster or on 
GPUs, environments for which debugging tools are primitive at best.
%  Finding these bugs required either multiple {\tt gdb} sessions 
%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
%GPUs.  
To locate the application bugs causing these illegal accesses, we initially added 
dynamic checks on  all region accesses for both CPUs and GPUs.  
Unsurprisingly, these checks added considerable runtime overhead.  
To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
We then rewrote all of our applications in this language and used our type checker to verify them.  Once the applications type checked we were
able to safely elide all the dynamic region access checks.

@@

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
component of each bar shows the overhead added by the dynamic checks.  In 
each figure the problem size stays the same as the number of processors increases
(strong scaling).  Figure~\ref{fig:amr_overhead} includes multiple problem sizes to show
how overhead is affected by changing problem size (weak scaling).  For cases where there
is an existing implementation to compare against we have included a dotted line indicating
baseline performance.  In a few cases (Figures~\ref{fig:fluid_overhead} and 
\ref{fig:amr4096}) checking overhead is the difference
between better and worse performance than the baseline.  For an explanation of overall
performance relative to the baseline implementations see \cite{Legion12}.

>>>>

We evaluate the design of Legion's static and dynamic semantics on four criteria:
expressivity (can real applications be written---Section~\ref{subsec:expressivity}),
overhead  (what are the dynamic checking costs---Section~\ref{subsec:overhead}),
scalability (can it enable hierarchical scheduling---Section~\ref{subsec:scalability}),
and performance  (does the performance increase from relaxed coherence modes 
warrant the increased semantic complexity---Section~\ref{subsec:relaxedperf}).
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language of Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of Keeneland
consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

\subsection{Expressivity}
\label{subsec:expressivity}
We evaluate Legion on three real-world applications.  To qualitatively gauge the 
expressivity of Legion, we introduce these applications
by describing features used in their implementations.
The Circuit example was already covered in detail in Section~\ref{sec:example}.

@@

The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with sub-refinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every level of refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one describes pieces at 
each level of refinement, and another describes relationships between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

@@

Dynamically creating and partitioning regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  
% FIXME
% I've deemphasized the point about allocate-then-partition vs. partition-then-allocate.
% If we want to put it back in it needs to be go in multiple places, not just here.
%Legion is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence 
are needed in some application;  region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multi-colorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications.
% written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

\subsection{Checking Overhead}
\label{subsec:overhead}
The first Legion implementation consisted of a C++ library of Legion primitives
\cite{Legion12} with no checking of region memory accesses.  
When testing applications on this system we frequently encountered memory corruption due to
illegal region accesses caused by application bugs.
In many cases, this corruption occurred between nodes in the cluster or on 
GPUs, environments for which debugging tools are primitive at best.
%  Finding these bugs required either multiple {\tt gdb} sessions 
%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
%GPUs.  
To locate the application bugs causing these illegal accesses, we initially added 
dynamic checks on  all region accesses for both CPUs and GPUs which  
added considerable runtime overhead.  
To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
We then rewrote the applications in this language and type checked them, at which point the
dynamic region access checks could be safely elided.

@@

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
component of each bar shows the overhead added by the dynamic checks.  In 
each figure the problem size stays the same as the number of processors increases
(strong scaling).  Figure~\ref{fig:amr_overhead} includes multiple problem sizes to show
how overhead is affected by changing problem size (weak scaling).  For cases where there
is an existing implementation to compare against we have included a dotted line indicating
baseline performance.  In a few cases (Figures~\ref{fig:fluid_overhead} and 
\ref{fig:amr4096}) checking overhead is the difference
between better and worse performance than the baseline.  The overall
performance relative to the baseline implementations is discussed in \cite{Legion12}.

### 2012-11-02 doc/types-pldi2013/related.tex
end

There have been many type and effect systems for ownership types
\cite{Boyapati03} including ones that leverage nested regions for describing
relationships (e.g., \cite{Clarke02}).  However, ownership type and effect systems
are primarily used for reasoning about determinism in object oriented languages and
don't capture the range of disjointness properties that can be specified in Legion.
Reasoning about disjoint data is the strong suit of separation logic \cite{Reynolds02}.  
%Concurrent separation logic\cite{Brookes04} has been 
%used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
%a mechanism for reasoning about independence\cite{Hayman06}.
While we have borrowed some separation logic notation, we ultimately chose to use a 
privileges system as our primary formalism because separation logic does not easily
support reasoning about the interleaving of operations to aliased regions of memory.

>>>>

There have been many type and effect systems for ownership types
\cite{Boyapati03} including ones that leverage nested regions for describing
relationships (e.g., \cite{Clarke02}).  However, ownership type and effect systems
are primarily used for reasoning about determinism in object oriented languages and
don't capture the range of disjointness properties in Legion.
Reasoning about disjoint data is the strong suit of separation logic \cite{Reynolds02}.  
%Concurrent separation logic\cite{Brookes04} has been 
%used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
%a mechanism for reasoning about independence\cite{Hayman06}.
While we have borrowed some separation logic notation, we chose to use a 
privileges system because separation logic does not easily
support reasoning about the interleaving of operations to aliased regions of memory.
